---
title: "in class assignment 2"
author: "Yang Xinrui"
date: 
output: 
  html_document: default
  pdf_document: default
  md_document: default
knit: (function(inputFile, encoding){rmarkdown::render(inputFile, encoding = encoding, output_format = "all", knit_root_dir = getwd())})
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
```


```{r}
d <- read_csv("_DataPublic_/vdem/1984_2022/vdem_1984_2022_external.csv")
```


# 1. Codebook lookup.
## i. What indicators regarding the quality of education are available in the V-Dem datasets? 
Education 15+ (e_peaveduc) ; Educational inequality, Gini (e_peedgini)

## ii. What are the data’s coverage (i.e., for which countries and years do we have data?)

```{r}
# countries
d |> select(country_name) |> distinct()
```

```{r}
# years
d |> select(year) |> distinct()
```

## iii. What are their sources? Provide the link to least 1 source.
Clio Infra (clio-infra.eu)


# 2. Subset by columns
## i. Create a dataset containing only the country-year identifiers and indicators of education quality.

```{r}
d_edu <- d |> select(country_name, country_id, year, e_peaveduc, e_peedgini) |> distinct()

d_edu
```

## ii. Rename the columns of education quality to make them informative.

```{r}
d_edu |>
  rename("Education" = "e_peaveduc", "Gini" = "e_peedgini",
         "Country" = "country_name", "ID" = "country_id",
         "Year" = "year")


d_edu <- d_edu |>
  rename("Education" = "e_peaveduc", "Gini" = "e_peedgini",
         "Country" = "country_name", "ID" = "country_id",
         "Year" = "year")

d_edu
```

# 3. Subset by rows 
## i. List 5 countries-years that have the highest education level among its population.
```{r}
d_edu |> slice_max(order_by = Education, n =5)
```

## ii. List 5 countries-years that suffer from the most severe inequality in education.

```{r}
d_edu |> slice_min(order_by = Gini, n =5)
```

# 4. Summarize the data
## i. Check data availability: For which countries and years are the indicators of education quality available?
```{r}
d_edu |>
  mutate(GDP_missing = is.na(Education), .after = Education) |>
  group_by(Country) |>
  summarise(N_GDP_missing = sum(GDP_missing))
```

```{r}
d_edu |>
  mutate(GDP_missing = is.na(Gini), .after = Gini) |>
  group_by(Country) |>
  summarise(N_GDP_missing = sum(GDP_missing))
```

## ii. Create two types of country-level indicators of education quality
### a. Average level of education quality from 1984 to 2022
```{r}
d_edu |> 
  group_by(Country) |>
  summarise(Education_average = mean(Education, na.rm = TRUE)) |>
  arrange(Country)
```
### b. Change of education quality from 1984 to 2022
```{r}
d_edu |> 
  filter(Year >= 1984 & Year <= 2022) |>
  group_by(Country) |>
  arrange(Year) |>
  summarise(Education_growth_2022_1984 = (last(Education) - first(Education)) / first(Education)) |>
  ungroup() |>
  arrange(Country)
```
## iii. Examine the data and briefly discuss: Which countries perform the best and the worst in terms of education quality in the past four decades?
```{r}
d_edu |> 
  group_by(Country) |>
  summarise(Education_average = mean(Education, na.rm = TRUE)) |>
  arrange(Education_average)
```

```{r}
d_edu |> 
  group_by(Country) |>
  summarise(Education_average = mean(Education, na.rm = TRUE)) |>
  arrange(desc(Education_average))
```

```{r}
d_edu |> 
  filter(Year >= 1984 & Year <= 2022) |>
  group_by(Country) |>
  arrange(Year) |>
  summarise(Education_growth_2022_1984 = (last(Education) - first(Education)) / first(Education)) |>
  ungroup() |>
  arrange(Education_growth_2022_1984)
```

```{r}
d_edu |> 
  filter(Year >= 1984 & Year <= 2022) |>
  group_by(Country) |>
  arrange(Year) |>
  summarise(Education_growth_2022_1984 = (last(Education) - first(Education)) / first(Education)) |>
  ungroup() |>
  arrange(desc(Education_growth_2022_1984))
```










